Cooperative Kernels: Safe Multitasking for Blocking Algorithms on GPUs

### 2016-11-10 write-ups/PLDI/paper.tex
Progress.

In effect, thread 0 of the spawning workgroup transmits the relevant
portion of its local state to the new threads.  We discuss in
Section~\ref{sec:semanticalternatives} the rationale for having all
new threads obtain this local state portion from thread 0, as well as
our reasons for selecting which variables to transmit via annotations
rather than transmitting the entire thread state.

@@

Many irregular algorithms require synchronisation across workgroups.
This can be achieved via a \emph{global barrier}, but a global barrier
is not provided as a primitive in OpenCL, CUDA or HSA.  This is
because, as discussed in Section~\ref{sec:intro}, an inter-workgroup barrier
requires fair scheduling of workgroups.

Because workgroups of a cooperative kernel are fairly scheduled, a
global barrier primitive can be provided.  We specify two such primitives: $\globalbarrier()$
and $\resizingglobalbarrier()$.

A $\globalbarrier()$ is a kernel-level function: if it appears in
conditional code then it must be reached by \emph{all} threads
executing the cooperative kernel.  On reaching a $\globalbarrier()$, a
thread waits until all threads have arrived at the barrier.  Once all
threads have arrived, the threads may proceed past the barrier with
the guarantee that all global memory accesses issued before the
barrier have completed.

A $\resizingglobalbarrier()$ is identical to a global barrier, except
that, semantically, every workgroup executes $\offerkill$ on reaching
the barrier, and every workgroup executes $\offerfork$ on leaving the
barrier.  Thus $\resizingglobalbarrier()$ is semantically equivalent to:
%
\lstset{basicstyle=\tt}
\begin{lstlisting}
  $\offerkill()$;
  $\globalbarrier()$;
  $\offerfork()$;
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt}

The $\globalbarrier()$ primitive can be implemented directly using the
atomic operations provided by the OpenCL 2.0 memory model~\cite{...},
but the implementation is involved, hence why we include this function
as a primitive.  As noted above, the $\resizingglobalbarrier()$
primitive is, semantically, syntactic sugar for a sequence of other
primitive calls.  We include it as a primitive because (a) we have
found that global barriers often provide a natural point for
cooperation with the scheduler, and (b) a custom implementation of
$\resizingglobalbarrier()$ that provides the semantics described
above can be made significantly more efficient than the above sequence of
calls.  We discuss our efficient implementation of
$\resizingglobalbarrier$ in detail in Section~\ref{sec:resizingbarrier}.

@@

\ADComment{TODO}

@@

\ADComment{Allowing zero to be killed.}

>>>>

In effect, thread 0 of the forking workgroup transmits the relevant
portion of its local state to the threads of the forked workgroups.
We discuss in Section~\ref{sec:semanticalternatives} the rationale for
having all new threads obtain this local state portion from thread 0,
as well as our reasons for selecting which variables to transmit via
annotations rather than transmitting the entire thread state.

@@

\ADComment{Maybe refer back to motivating examples?}
Many irregular algorithms require synchronisation across workgroups.
This can be achieved via a \emph{global barrier}, but a global barrier
is not provided as a primitive in OpenCL, CUDA or HSA.  This is
because, as discussed in Section~\ref{sec:intro}, an inter-workgroup barrier
requires fair scheduling of workgroups.

Because workgroups of a cooperative kernel are fairly scheduled, a
global barrier primitive can be provided.  We specify two such primitives: $\globalbarrier$
and $\resizingglobalbarrier$.

Our $\globalbarrier$ primitive is a kernel-level function: if it appears in
conditional code then it must be reached by \emph{all} threads
executing the cooperative kernel.  On reaching a $\globalbarrier()$, a
thread waits until all threads have arrived at the barrier.  Once all
threads have arrived, the threads may proceed past the barrier with
the guarantee that all global memory accesses issued before the
barrier have completed.

The $\resizingglobalbarrier$ primitive is also a kernel-level
function.  It is identical to $\globalbarrier$, except that it caters
for cooperation with the scheduler: by issuing a
$\resizingglobalbarrier$ the programmer indicates that the cooperative
kernel is prepared to proceed after the barrier with fewer workgroups,
or with additional workgroups joining the computation.
Semantically, it is as if every workgroup executes $\offerkill$ on reaching the
barrier, and every workgroup executes $\offerfork$ on leaving the
barrier.  Thus $\resizingglobalbarrier()$ is semantically equivalent
to: \ADComment{Does the order of these three calls matter, semantically?}
%
\lstset{basicstyle=\tt}
\begin{lstlisting}
  $\offerkill()$;
  $\globalbarrier()$;
  $\offerfork()$;
\end{lstlisting}
\lstset{basicstyle=\scriptsize\tt}

The $\globalbarrier$ primitive can be implemented following any
inter-workgroup barrier design, e.g.~\cite{...}, and the atomic
operations provided by the OpenCL 2.0 memory model enable a
memory-safe implementation~\cite{...}.  However, implementing such a
barrier is involved, hence why we include this function as a
primitive.  As noted above, the $\resizingglobalbarrier$ primitive
is, semantically, syntactic sugar for a sequence of other primitive
calls.  We include it as a primitive because (a) we have found that
global barriers often provide a natural point for cooperation with the
scheduler, and (b) a custom implementation of
$\resizingglobalbarrier()$ that provides the semantics described above
can be made significantly more efficient than the above sequence of
calls.  We discuss our efficient implementation of
$\resizingglobalbarrier$ in detail in
Section~\ref{sec:resizingbarrier}.

@@

\ADComment{TODO.  We want to show the examples from Section~\ref{sec:background} ported to use cooperative kernels.  We want to make it clear that it is easy to place the primitives.}

@@

\ADComment{Allowing zero to be killed; the following is some text we had before:}
